taxadb: A High-Performance Local Taxonomic Database Interface

### 2019-06-28 paper/manuscript.Rmd

---
title: "taxadb: A High-Performance Local Taxonomic Database Interface"
author:
  - name: "Kari Norman"
    affiliation: ucb
  - name: "Jorrit Poelen"
    affiliation: jorritt
  - name: "Scott Chamberlain"
    affiliation: ropensci      
  - name: "Carl Boettiger"
    affiliation: ucb, 1
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: jorrit
    address: "Independent consultant, Oakland, CA, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA"
abstract: |
  A familiar and growing challenge in ecological and evolutionary research is that of reconciling scientific names of relevant taxa when combining data from separate sources. While this problem is already well understood and numerous naming authorities have been created to address the issue, most researchers lack a fast, consistent, and intuitive way to reconcile taxonomic names.  Here, we present `taxadb` R package to address this gap.  In contrast to existing tools, `taxadb` provides the following:  
  1) `taxadb` accesses established naming authorities to resolve synonyms, IDs, and hiearchical classification.  
  2) `taxadb` creates a local database, managed automatically from within R, that provides fast operations on millions of taxonomic names.  
  3) `taxadb` provides a consistent and intuitive data format.  
  4) `taxadb` is built on a simple, extensible and language agnostic design that can easily accomodiate new authorities.

@@

```{r message=FALSE, include = FALSE}
library(kableExtra)
library(magrittr)
library(tidyverse)
library(taxadb)
library(printr)
pkgconfig::set_config("dplyr::na_matches" = "never")

@@

After installing the `taxadb` R package, users can create local copies of data from any of the providers using the `td_create()` function and specifying the provider abbreviation (see Table 1), or `all` to install all available providers:

@@

@@

<!--  Next 2 paragraphs should possibly be left to a manual, but are likely to be overlooked aspects there... -->
This one-time download will download and import local copies of the provider data. By default `taxadb` creates a MonetDBLite database instance [@monetdblite], a columnar-oriented relational database requiring no additional installation while also providing persistent disk-based storage, and data is stored in the appropriate location specified for applications by the operating sytstem [@rappdirs]. Users can customize this location, or opt for any alternative relational database backend by providing alternative arguments, though the default MonetDBLite system can be an order of magnitude faster than popular alternatives such as Postgres [@monetdblite].  `taxadb` will automatically handle opening, caching, and closing the MonetDBLite connection, but note that one limitation of MonetDBLite is the restriction against concurrent access: two separate R sessions on the same machine cannot both access the database at the same time. Users who need concurrent access from multiple sessions may wish to use a standalone MonetDB server or other database server instead.

All `taxadb` functions can also operate in one-time-use mode without first installing a standalone server.  If a function requests a provider which has not been installed, the relevant table can be downloaded and read directly into memory.  This requires sufficient memory be available to hold the entire table (typically 1-2 GB) and will only be cached for the R session.  This can be useful in special cases such as execution on remote servers where this will be faster than importing the data locally, or to allow other packages to use methods such as `get_ids()` internally without assuming users will have first created a local copy with `td_create()`. Specify `db=NULL` on any `taxadb` function to force this in-memory dispatch.

@@

```{r}
trait_data <- read_tsv(system.file("extdata", "trait_data.tsv", package="taxadb"))
status_data <- read_tsv(system.file("extdata", "status_data.tsv", package="taxadb"))
```

>>>>

---
title: "taxadb: A High-Performance Local Taxonomic Database Interface"
author:
  - name: "Kari Norman"
    affiliation: ucb
#  - name: "Jorrit Poelen"
#    affiliation: jorritt
  - name: "Scott Chamberlain"
    affiliation: ropensci      
  - name: "Carl Boettiger"
    affiliation: ucb, 1
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: jorrit
    address: "Independent consultant, Oakland, CA, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA"
abstract: |
  A familiar and growing challenge in ecological and evolutionary research is that of reconciling scientific names of relevant taxa when combining data from separate sources. While this problem is already well understood and numerous naming authorities have been created to address the issue, most researchers lack a fast, consistent, and intuitive way to reconcile taxonomic names.  Here, we present `taxadb` R package to address this gap.  In contrast to existing tools, `taxadb` provides the following:  
  1) `taxadb` accesses established naming authorities to resolve synonyms, IDs, and hiearchical classification.  
  2) `taxadb` creates a local database, managed automatically from within R, that provides fast operations on millions of taxonomic names.  
  3) `taxadb` provides a consistent and intuitive data format.  
  4) `taxadb` is built on a simple, extensible and language agnostic design that can easily accomodiate new authorities.

@@

```{r message=FALSE, include = FALSE}
library(kableExtra)
library(magrittr)
library(tidyverse)
library(taxadb)
library(printr)

## see https://blog.rstudio.com/2017/06/13/dplyr-0-7-0/, 
## https://community.rstudio.com/t/why-does-na-match-na-when-joining-two-dataframes/28785/3
pkgconfig::set_config("dplyr::na_matches" = "never")

@@

<!--
### Installation

`taxadb` can be installed from GitHub using the `remotes` R package:

```{r eval=FALSE}
remotes::install_github("cboettig/taxadb")
```
-->

@@

`taxadb` functions will automatically set up a local, persistent database if one has not yet been created.

```{r}
get_ids("Gadus morhua")
```

This one-time setup will download, uncompress, and import the data into persistant storage (using the appropriate location specified by the operating sytstem [see @rappdirs], or configured using the environmental variable `TAXADB_HOME`).  The example above searches for names in ITIS, the default provider, which can be configured using the `provider` argument. Any future function calls to this function or any other function using data from the same names provider will be able to access this data rapidly without the need for processing or an internet connection.

Users can also explicitly trigger this one-time setup using `td_create()` and specifying the provider abbreviation (see Table 1), or simply using `all` to install all available providers:

@@

`taxadb` functions like `get_ids()` and `td_create()` take an optional argument, `db`, to an external database connection.  `taxadb` will work with most DBI-compliant databases such as MySQL or Postgres [see @DBI], but will be much faster when using a column-oriented database engine such as `duckdb` or `MonetDBLite`.  These latter options are also much easier for most users, since each can be installed directly as an R package. `taxadb` will default to the fastest available option.  `taxadb` can also run without a database backend by setting `db=NULL`, though some functions will require a lot (2-20 GB) of free RAM for this to work with many of the large names providers.

<!--
`taxadb` does not require the use of an external database system, but will automatically make use of the fastest database server available. `taxadb` can also use any DBI compliant database accessible from R, such as MySQL or Postgres [see @DBI], though the default `duckdb` or `MonetDBLite` will be both significantly faster and easier to use.  
-->

<!--  Next 2 paragraphs should possibly be left to a manual, but are likely to be overlooked aspects there... -->
By default `taxadb` creates a MonetDBLite database instance [@monetdblite], a columnar-oriented relational database requiring no additional installation while also providing persistent disk-based storage. `taxadb` will automatically handle opening, caching, and closing the database connection.

<!-- 
now sidestepped

but note that one limitation of MonetDBLite is the restriction against concurrent access: two separate R sessions on the same machine cannot both access the database at the same time. Users who need concurrent access from multiple sessions may wish to use a standalone MonetDB server or other database server instead.
-->

It is not necessary to call `td_create()`.  Specify `db=NULL` on any `taxadb` function to force this in-memory dispatch.

@@

```{r message=FALSE}
trait_data <- read_tsv(system.file("extdata", "trait_data.tsv", package="taxadb"))
status_data <- read_tsv(system.file("extdata", "status_data.tsv", package="taxadb"))
```
